Outlier Detection and Comparison of Origin-Destination Flows using Data Depth

### 2018-05-31 GIScience_2018/giscience_r2/giscience_r2/giscience2018_origional.tex
Jun's version

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author{Myeong-Hun Jeong}{Department of Civil Engineering, Chosun University, Gwangju, Republic of Korea}{mhjeong@chosun.ac.kr}{[orcid]}{[funding]}
\author{Junjun Yin}{Social Science Research Institute; Institute for CyberScience, Penn State University, PA, USA}{jyin@psu.edu}{[0000-0002-4196-2439]}{[This work used the Extreme Science and Engineering Discovery Environment (XSEDE), which is supported by National Science Foundation grant number ACI-1548562]}
\author{Shaowen Wang}{Department of Geography and Geographic Information Science, University of Illinois at Urbana-Champaign, IL, USA}{shaowen@illinois.edu}{[orcid]}{[funding]}

@@

\begin{abstract}
Advances in location-aware technology have lead to the generation of a huge volume of trajectory data.
Origin-destination (OD) trajectories provide rich information on urban flow and transport demand.
This study presents a new methodology to detect OD flow outliers and conduct hypothesis testing between two OD flows in terms of the variations of spatial extent, namely, spread.
The proposed method is based on data depth, which measures the centrality and outlyingness of a point with respect to a given dataset in $\mathbb{R}^d$.
Based on the center-outward ordering property, the proposed method analyzes the underlying characteristics of OD flows, such as location, outlyingness, and spread.
The ability of the proposed method to detect OD anomalies is compared with that of the Mahalanobis distance approach, and an F-test is used to verify the difference in scale. Empirical evaluation has demonstrated that the proposed method effectively identifies OD flows outliers in an interactive way. Furthermore, the proposed method can provide new perspectives by considering the overall structure of data when comparing two different OD flows in terms of scale.

@@

It is worth noting that flow mapping approaches frequently suffer from the modifiable areal unit problem (MAUP). 
Essentially, MAUP is the influence of different aggregations determined by location on the presentation of coherent patterns.
Kernel-based flow estimation and smoothing are used to overcome different spatial resolutions \cite{guo14IEEETVCG}. 
Instead of attempting to find the best areal unit by which to partition urban space and aggregates the OD flows, this study adopted the established traffic analysis zones of New York City as base unit. That said, the proposed method can be adapted to other implementation of areal units. In this study, New York City taxi trip data includes the origin and destination within traffic analysis zones, while ignoring the intermediate locations of the actual routes.
It is not necessary to reconstruct individual movements for flow estimation (see \cite{duckham16ICGIS}).

@@

The bag plot presents the data using the following attributes:location is represented by the depth median; spread or the spatial extent of bag; correlation or the orientation of the bag; and skewness, as represented by the shape of the bag and the fence \cite{rousseeuw99AS}.
In Figure \ref{fig:OD_0521}, we observe that some forward OD flows have higher counts than their paired reverse OD flows. We also note the relatively linear correlation between forward OD flows and reverse OD flows and the skewness of forward (reverse) OD flows.

It is also possible to detect the outliers of OD flows of two different time stamps.
In Figure \ref{fig:OD_0721_0521}, we visualize the OD flows recorded on two separated days.
Comparing the two sets of OD flows not only indicates the central region of OD flows, it also distinguishes the significantly different OD flows.

@@

For data cleaning process, this study used Hadoop with Pig.
We developed a Hadoop program to handle the large data volume, which was composed of 173 million taxi trip records, remove trips with invalid OD coordinates, and assign each OD locations into the corresponding traffic analysis zone.
To implement the OD flow outlier detection, this study used R.
The computing environment used Amazon Web Service and the Bridges supercomputer at the Pittsburgh Supercomputing Center. 
This study only evaluated OD flows more than 10 trips, as the low trip number OD flows could have distorted the view of the data.
All the code will be released as open source (the link to the code will added later and currently is available upon request).

@@

As an alternative method, the state-of-the-art Mahalanobis distance-based approach detected similar outlier OD flows.
However, the number of outliers detected was different.
This occurred because our OD flows data had heavy tail distributions, that is, many of the OD flows with a long distance from the depth median depicted in Figure \ref{fig:weekdays_rare}. 
Mahalanobis distance is known to be inadequate when the underlying data have heavy tailed distributions \cite{wilcox12Book}. 
Thus, the presence of outliers may mask the detection of other outliers in Mahalanobis distance approach.
Furthermore, it can only detect OD flow outliers with high numbers of trips during time $t_1$ and time $t_2$. 
It is difficult to detect OD flows outliers that have different properties, such as substantial differences in the number of trips when comparing with time $t_1$ and time $t_2$.

@@

This paper provides a new methodology for identifying outlier OD flows and the difference in scale between two different OD flows at $t_1$ and $t_2$.
The proposed method is based on the concept of data depth.
Data depth is robust statistics, which is suited to non-Gaussian distribution of the underlying datasets.
Compared with standard statistics, this approach enhances understanding of the differences and the magnitude of the differences between two OD flows.

@@

This study made no attempt to incorporate geographic contexts such as locational circumstances or surrounding environment in understanding OD flows. 
Further investigations will focus on integrating the analysis of OD flows with geographic context.
Such an effort will lead to knowledge discovery and better understandings regarding the dynamics of urban flow.

>>>>

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author{Myeong-Hun Jeong}{Department of Civil Engineering, Chosun University, Gwangju, Republic of Korea}{mhjeong@chosun.ac.kr}{[orcid]}{[funding]}
\author{Junjun Yin}{Social Science Research Institute; Institute for CyberScience, Penn State University, PA, USA}{jyin@psu.edu}{[orcid]}{[funding]}
\author{Shaowen Wang}{Department of Geography and Geographic Information Science, University of Illinois at Urbana-Champaign, IL, USA}{shaowen@illinois.edu}{[orcid]}{[funding]}

@@

\begin{abstract}
Advances in location-aware technology have resulted in the generation of  a huge volume of trajectory data. Origin-destination (OD) trajectories provide rich information on urban flow and transport demand. This study presents a new methodology to detect OD flows outliers and conduct hypothesis testing between two OD flows in terms of the variations of spatial extent, that is, spread. The proposed method is based on data depth, which measures the centrality and outlyingness of a point with respect to a given dataset in $\mathbb{R}^d$. Based on the center-outward ordering property, the proposed method analyzes the underlying characteristics of OD flows, such as location, outlyingness, and  spread. The ability of the proposed method to detect OD anomalies is compared with that of the Mahalanobis distance approach, and an F-test is used to verify the difference in scale. Empirical evaluation has demonstrated that the proposed method effectively identifies  OD flows outliers in an interactive way. Furthermore, the proposed method  can provide new perspectives by considering the overall structure of data when comparing two different OD flows in terms of scale.

@@

It is worth noting that flow mapping approaches frequently suffer from the modifiable areal unit problem (MAUP).Essentially, MAUP is the influence of different aggregations determined by location on the presentation of coherent patterns.  Kernel-based flow estimation and smoothing are used to overcome different spatial resolutions \cite{guo14IEEETVCG}. 
Instead of attempting to find the best areal unit by which to partition urban space and aggregates the OD flows, this study adopted the established traffic analysis  zones of New York City as base unit. That said, the proposed method can be adapted to other implementation of areal units. In this study, New York City taxi trip data includes the origin and destination within traffic analysis zones, while ignoring the intermediate locations of the actual routes.
It is not necessary to reconstruct individual movements for flow estimation (see \cite{duckham16ICGIS}).

@@

The bag plot presents the data using the following attributes: location is represented by the depth median; spread or the spatial extent of bag; correlation or the orientation of the bag; and skewness, as represented by the shape of the bag and the fence \cite{rousseeuw99AS}.  In Figure \ref{fig:OD_0521}, we observe that some forward OD flows have higher counts than their paired reverse OD flows. We also note the relatively linear correlation between forward OD flows and reverse OD flows and the skewness of forward (reverse) OD flows.

It is also possible to detect the outliers of OD flows of two different time stamps. In Figure \ref{fig:OD_0721_0521}, we visualize the OD flows recorded on two different days.
Comparing the two sets of OD flows not only indicates the central region of OD flows, it also distinguishes the significantly  different OD flows.

@@

For data cleaning process, this study used Hadoop with Pig. We developed a Hadoop program to handle the large volume of data, which was composed of 173 million taxi trip records, remove trips with invalid OD coordinates, and assign each OD locations into the corresponding traffic analysis zone.  
To implement the OD flow outlier detection, this study used R. The computing environment used Amazon Web Service and the Bridges supercomputer at the Pittsburgh Supercomputing Center. This study only evaluated OD flows more than 10 trips, as the low trip number OD flows could have distorted the view of the data. All the code will be released as open source (the link to the code will added later and currently is available upon request).

@@

As an alternative method, the state-of-the-art Mahalanobis distance approach  detected similar outlier OD flows. However, the number of outliers detected was different. This occurred because our OD flows data had heavy tail distributions, that is, many of the OD flows with a long distance from the depth median depicted in Figure \ref{fig:weekdays_rare}. Mahalanobis distance is known to be inadequate when the underlying data have heavy tail distributions \cite{wilcox12Book}. Thus, the presence of outliers may mask the detection of other outliers in Mahalanobis distance approach. Furthermore, it can only detect OD flow outliers with high numbers of trips during time $t_1$ and time $t_2$. It is difficult to detect OD flows outliers that have different properties, such as substantial differences in the number of trips when comparing with time $t_1$ and time $t_2$.

@@

This paper provides a new methodology for identifying outlier OD flows and the difference in scale between two different OD flows at $t_1$ and $t_2$. The proposed method is based on the concept of data depth. Data depth is robust statistics, which is suited to non-Gaussian distribution of the underlying datasets. Compared with standard statistics, this approach enhances understanding of the differences and the magnitude of the differences between tow OD flows.

@@

This study made no attempt to consider geographic contexts such as locational circumstances or surrounding environment in understanding OD flows. Ultimately, further investigation should focus on integrating the analysis of OD flows with geographic context. Such an effort will lead to knowledge discovery and understanding the dynamics of urban flow.
